---
title: "Class 14: RNASeq mini project"
author: "Nicole (PID: A18116280)"
format: pdf
toc: true
---

## Background 

Here we work through a complete RNASeq analysis project. The input data comes from a knock-down experiment of a HDX gene. 
## Data Import

Reading the `count()` and `metadata` CSV files 


```{r}
counts <- read.csv("GSE37704_featurecounts (1).csv", row.names = 1)
metadata <- read.csv("GSE37704_metadata.csv")
```

Check on data structure

```{r}
head(counts)
```
```{r}
metadata
```


```{r}
head(metadata)
```

Some book-keeping is required as there looks to be a mis-match between metadata and counts columns

```{r}
ncol(counts)
```
```{r}
nrow(metadata)
```

Looks like we need to get rid of the first "length" column of our `counts` object. 

```{r}
cleancounts <- counts[ , -1]
```

```{r}
colnames(cleancounts)
```

```{r}
metadata$id
```

```{r}
all( colnames(cleancounts) == metadata$id)
```

### Remove zero count genes

There are lots of genes with zero counts. We can remove these from further analysis
```{r}
head(cleancounts)
```

```{r}
to.keep.inds <- rowSums(cleancounts) > 0
nonzero_counts <- cleancounts[to.keep.inds,]
```

## DESeq analysis
Load the package

```{r, message=FALSE}
library(DESeq2)
```

Setup DESeq object

```{r}
dds <- DESeqDataSetFromMatrix(countData = nonzero_counts,
                              colData = metadata,
                              design = ~condition)
```

Run DESeq

```{r}
dds <-DESeq(dds)
```

get results

```{r}
res <- results(dds)
head(res)
```

## Data Visualization

```{r}
library(ggplot2)

ggplot(res) +
  aes(log2FoldChange, -log(padj) ) +
  geom_point()

```

Add threshold lines for fold-change and P-value and color our subset of genes that make these threshold cut-offs in the plot

```{r}
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2] <- "blue"
mycols[ res$padj > 0.05 ] <- "gray"

ggplot(res) +
  aes(log2FoldChange, -log(padj), color = mycols) +
  geom_point() +
  geom_vline(xintercept = c(-2, 2), color = "red") +
  geom_hline(yintercept = -log(0.05), color = "red") +
  scale_color_identity()
  
```

## Add Annotation

Add gene symbols and entrez ids

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(x=org.Hs.eg.db, 
                     keys=row.names(res),
                     keytype = "ENSEMBL",
                     column = "SYMBOL")
```
```{r}
res$symbol <- mapIds(x=org.Hs.eg.db, 
                     keys=row.names(res),
                     keytype = "ENSEMBL",
                     column = "ENTREZID" )
```


## Pathway Analysis

Run gage analysis

```{r, message=FALSE}
library(gage)
library(gageData)
library(pathview)
```

We need a named vector of fold-change values as input for gage

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
data(kegg.sets.hs)

keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
head(keggres$less, 2)
```
```{r}
pathview(pathway.id = "hsa04110", gene.data=foldchanges)
```

![](hsa04110.xml)


### GO terms

Same analysis but using GO genesets rather than KEGG
```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets)
```

```{r}
head(gobpres$less, 4)
```

### Reactome

Lots of folks like the reactome web interface. You can also run this as an R function but lets look at the website first < https://reactome.org/ >

The website wants a text file with one gene symbol per line of the genes you want to map to pathways. 

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), ]$symbol
head(sig_genes) #res$symbol
```

and write out to a file: 
```{r}
write.table(sig_genes, file="significant_genes.txt", 
            row.names=FALSE, col.names=FALSE, quote=FALSE)
```



## Save Our Results

```{r}
write.csv(res, file="myresults.csv")
```


