---
title: "Class 5: Data Viz with ggplot"
author: "Nicole (A18116280)"
format: pdf
---

Today we are exploring the **ggplot** package and how to make nice figures in R. 

There are lots of ways to make figures and plots in R. These include: 

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot. 

```{r}
head(cars)
```

We can simply pass to the `plot()` function. 

```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice looking in some folks eyes. 

Let's see how we can plot this with **ggplot2**...

1st I need to install this add-on package. For this we use the `install.package()` function - **WE DO THIS IN THE CONSOLE, NOT our report**. This is a one time only deal. 

2nd we need to load the package with the `library()` functino every time we want to use it. 

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

- **data** (i.e a data.frame with the things you want to plot), 
- aesthitics **aes()** that map the columns of data to your plot features (i.e. aesthitics)
- geoms like **geom_point()** that srt how the plot appears

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

```{r}
hist(cars$speed)
```

> For simple "canned" graphs base R is quicker but as things get more custom and elaborate then ggplot wins out...

Let's add more layers to our ggplot

Add a line showing the relationship between x and y
Add a title
Add custom axis labels "Speed (MPH)" and "Distance (ft)"
Change the theme...

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() + 
  geom_smooth(method=lm, se=FALSE) +
  labs(title="Silly plot of Speed vs Stoping distance", 
       x="Speed (MPH)", 
       y="Distance (ft)") + 
  theme_bw()
```

## Going further

Read some gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)

head(genes)
```

> Q1. How many genes are in this wee dataset?

```{r}
nrow(genes)
```


> Q2. How many "up" regulated genes are there?

```{r}
sum( genes$State == "up" )
```

A useful function for counting up occurrences of things in a vector is the `table()` function. 

```{r}
table(genes$State)
```

Make a v1 figure

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, 
      y=Condition2, 
      col=State) +
  geom_point() 

p
```

```{r}
p + scale_colour_manual( values=c("blue" , "gray", "red")) +
  labs(title="Expression changes upon drug treatment", 
      x="Control (no drug)", 
      y="Treatment (with Drug)") +
theme_bw()
```

## More Plotting 

Read in the gapminder dataset
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Lets have a wee peak

```{r}
head( gapminder, 3)
```

```{r}
tail( gapminder, 3)
```

> Q4. How many different country values are in this dataset?

```{r}
nrow(gapminder)
```

```{r}
length( table(gapminder$country) )
```

> Q5. How many different continent values are in this dataset?

```{r}
unique(gapminder$continent)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text()
```

I can use the **ggrepel** package to make more sensible lables here. 

```{r}
library(ggrepel)

ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text()
```

I want a separate pannel per continent 

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent, label=country) +
  geom_point() +
  facet_wrap(~continent)
```

## Summary

The main advantages of ggplot over base R are:

1. Layered Grammar: ggplot uses a consistent, layered approach—data, aesthetics, and geometry—making it easier to build and customize complex plots step by step, unlike base R which requires different functions and arguments for each plot type [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[6]](https://drive.google.com/file/d/19AkbGsbG0Rc8Nn_6wdQ8Fi_OFV9vcMIx/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

2. Publication Quality: ggplot produces attractive, publication-ready figures with sensible defaults, while base R plots often need more manual tweaking to look polished [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[6]](https://drive.google.com/file/d/19AkbGsbG0Rc8Nn_6wdQ8Fi_OFV9vcMIx/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

3. Reproducibility: ggplot code is modular and scriptable, making it easier to reproduce and update plots by changing data or layers [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[6]](https://drive.google.com/file/d/19AkbGsbG0Rc8Nn_6wdQ8Fi_OFV9vcMIx/view?usp=drivesdk).

4. Customization: Mapping data columns to visual features (color, size, shape) and adding legends, labels, and themes is straightforward in ggplot, but more cumbersome in base R [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[6]](https://drive.google.com/file/d/19AkbGsbG0Rc8Nn_6wdQ8Fi_OFV9vcMIx/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

5. Scalability: For simple plots, base R is quicker, but for complex, multi-layered figures, ggplot is more concise and manageable [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk).

Which of these advantages do you want to explore further?