---
title: "Class 7: Machine Learning 1"
author: "Nicole (PID:A18116280)"
format: pdf
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction. 

## K-means clustering

To see how this works let's first makeup some data to clister where we know what the answer should be. We can use the `rnorm()` function to help here:

```{r}
hist( rnorm(500, mean = 5) )
```

```{r}
x <- c( rnorm(30, mean = -3), rnorm(30, mean = 3) )
y <- rev(x)
```

```{r}
x <- cbind(x,y)
plot(x)
```

The function for K-means clustering in "base" R is `kmeans()`

```{r}
k <- kmeans(x, centers = 2)
k
```

To get at the results of the returned list object we can use the dollar `$` syntax

>Q. How many points are in each cluster?

```{r}
k$cluster
```

>Q. What 'component' of your result object details
  - cluster assignment/membership?
  - cluster center?
  
```{r}
k$centers
```

>Q. Make a clustering results figure of the data colored by cluster membership?

```{r}
plot(x, col = c("red", "blue"))
```

```{r}
plot(x, col = k$cluster, pch = 16 )
points(k$centers, col="blue", pch = 15, cex = 2)
```

K-means clustering is very popular as it is very fast and relatively straight forward: it takes numeric date as input and returns the cluster membership vector etc.

The "issue" is we tell `kmeans()` how many clusters we want!

>Q. Run kmeans again and cluster into 4 grps/clusters and plot the results like we did above?

```{r}
k4 <- kmeans(x, centers =4)
plot(x, col = k4$cluster)
points(k4$centers, pch =15)
```

Scree plot to pick k `centers` value

brute-force
```{r}
k1 <- kmeans(x, centers=1)
k2 <- kmeans(x, centers=2)
k3 <- kmeans(x, centers=3)
k4 <- kmeans(x, centers=4)
k5 <- kmeans(x, centers=5)
```

```{r}
z <- c(k1$tot.withinss, 
  k2$tot.withinss, 
  k3$tot.withinss, 
  k4$tot.withinss, 
  k5$tot.withinss)

plot(z, typ="b")
```

```{r}
n <- NULL
for (i in 1:5) {
  n <- c(n, kmeans(x, centers=i)$tot.withinss)
}

plot(n, typ="b")
```

##Hierarchical Clustering

The meain "base" R function for Hierarchical Clustering is called `hclust()`. Here we can't just input our data we need to first calculate a distance matrix for our data and use this as input to `hclust()`

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```
There is a plot method for hclust results lets try it

```{r}
plot(hc)
abline(h=8, col="red")
cutree(hc, h=8)
```

To get our cluster "membership" vector (i.e. our main clustering result) we can "cut" the tree at a given height or at a height that yields a given "k" groups. 

```{r}
cutree(hc, h=8)
```

```{r}
grps <- cutree(hc, k=2)
```

>Q. Plot the data with our hclust result coloring

```{r}
plot(x, col=grps)
```

#Principal Component Analysis (PCA)

## PCA if UK food data
Import food data from an online CVS file:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x

```


```{r}
x <- read.csv(url, row.names =1)
x
```

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), col=rainbow(nrow(x)))
```


There is one plot that can be useful for smalle dataset:

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

> Main point: It can be difficult to spot major trends and patterns even in relatively small multivariate datasets (here we only have 17 dimensions, typically we have 1000). 

## PCS to the rescu

The main function in "base" R for PCS is called `prcomp()`

I will take the transpose of our data so the "food" are in the columns:

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

```{r}
cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=cols, pch=16)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col=cols)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

PCA looks super useful and we will come back to describe this further next day :))))
